---
title: "Create Humboldt Extension with unique target scope per row"
author: 
  - name: "[Yi-Ming Gan](https://orcid.org/0000-0001-7087-2646)"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    df_print: paged
---

## Read data

The scope terms of Humboldt Extension here were populated with ` | ` pipe separated values. But they can be individual rows. The script here is to transfrom them into individual rows.

```{r, read-data, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)

# Get tsv files from Google Drive 
event <- read_tsv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTzxqpYCe1tVdichPPMCVgP9fyY6duJrtgyO8zGwm7xMKL5WLb3l6MPq0Ke5TIlwU97ovdZ__ptkkMw/pub?gid=0&single=true&output=tsv", col_names = TRUE, show_col_types = FALSE)
occ <- read_tsv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTzxqpYCe1tVdichPPMCVgP9fyY6duJrtgyO8zGwm7xMKL5WLb3l6MPq0Ke5TIlwU97ovdZ__ptkkMw/pub?gid=53360819&single=true&output=tsv", col_names = TRUE, show_col_types = FALSE)
humboldt <- read_tsv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTzxqpYCe1tVdichPPMCVgP9fyY6duJrtgyO8zGwm7xMKL5WLb3l6MPq0Ke5TIlwU97ovdZ__ptkkMw/pub?gid=604710631&single=true&output=tsv", col_names = TRUE, show_col_types = FALSE)
emof <- read_tsv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTzxqpYCe1tVdichPPMCVgP9fyY6duJrtgyO8zGwm7xMKL5WLb3l6MPq0Ke5TIlwU97ovdZ__ptkkMw/pub?gid=2088877587&single=true&output=tsv", col_names = TRUE, show_col_types = FALSE)
target <- read_tsv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTzxqpYCe1tVdichPPMCVgP9fyY6duJrtgyO8zGwm7xMKL5WLb3l6MPq0Ke5TIlwU97ovdZ__ptkkMw/pub?gid=872194191&single=true&output=tsv", col_names = TRUE, show_col_types = FALSE)

# show how target table looks like here
target
```

## Create target scope for each target taxon and target life stage combination

I am exploring if a humboldt table per target scope combination instead of piping the values makes sense.

```{r, create-target, message=FALSE, warning=FALSE}
record_target <- target %>% 
  # drop these columns because they are not part of humboldt terms
  select(-id, -scientificNameID, -taxonRank, -taxonRemarks) %>%
  # rename the field because this will be added into the humboldt table
  rename(targetTaxonomicScope = scientificName)

humboldt_ratified <- humboldt %>%
  # remove columns that are not needed
  select(-eventDuration, -targetTaxonomicScope, -targetLifestageScope, -excludedLifeStageScope) %>%
  # remove BROKE_WEST row because we do not want to join this record with target
  filter(eventID != "BROKE_WEST") %>%
  left_join(record_target, by = character()) %>%
  arrange(eventID)

# this step removes columns that were already excluded in humboldt_ratified table before appending the BROKE_WEST record back to the data frame
broke_west <- humboldt %>% 
  filter(eventID == "BROKE_WEST") %>%
  select(-eventDuration, -targetTaxonomicScope, -targetLifestageScope, -excludedLifeStageScope)

# append humboldt record eventID = BROKE_WEST back to the table
humboldt_ratified <- humboldt_ratified %>% 
  bind_rows(broke_west) %>%
  arrange(eventID)

# show how Humboldt table looks like for a single Event
humboldt_ratified %>% 
  filter(eventID == "BROKE_WEST_RMT_001") %>%
  select(eventID, targetTaxonomicScope, targetLifeStageScope)
```

## Populate is__ and has__ Humboldt terms programmatically

### isAbundanceReported

```{r, isAbundanceReported, message=FALSE, warning=FALSE}
humboldt_ratified <- humboldt_ratified %>% 
  mutate(
    isAbundanceReported = map2_lgl(eventID, targetTaxonomicScope, ~ {
      occurrence <- occ %>%
        filter(eventID == .x)
      # all targets are at taxonRank = family level 
      any(occurrence$family == .y &
      # preference of "all" for all life stages over a list of possible life stages because I do not know whether the list of targetLifeStageScope is a **complete** list of life stages for the taxon
            (.y == "all" | occurrence$lifeStage %in% targetLifeStageScope & !is.na(occurrence$individualCount)))
    })
  )
```

### absentTaxa

Is this only for taxa from `targetTaxonomicScope`? When to use this field and when to use Occurrence record with `occurrenceStatus` = `absent`? 

### isAbsenceReported

Is this for `absentTaxa` or dwc:Occurrence record with `occurrenceStatus` = `absent`?

### isTaxonomicScopeFullyReported

`true` for all Events.

### areNonTargetTaxaFullyReported 

`true` for all Events.

### hasNonTargetTaxa

I felt that it is a little confusing when the value of this term refers to Event level but each row in the Humboldt table was populated at target level scope (with unique combination of `targetTaxonomicScope` and `targetLifeStageScope`). `hasNonTargetTaxa` has to look at all rows within the Humboldt table with the same `eventID`. The humboldt table seems to conflate target scopes and Event.

```{r, hasNonTargetTaxa, message=FALSE, warning=FALSE}
# Get unique eventIDs from occ 
occ_eventIDs <- unique(occ$eventID)

humboldt_ratified <- humboldt_ratified %>%
  mutate(
    hasNonTargetTaxa = map_lgl(eventID, ~{
      if (.x %in% occ_eventIDs) {
        occ_families <- occ %>%
          filter(eventID == .x) %>%
          pull(family)
        any(!occ_families %in% targetTaxonomicScope[eventID == .x])
      } else {
        FALSE  # If eventID is not present in occ, return FALSE 
      }
    })
  )

humboldt_ratified %>% 
  filter(hasNonTargetTaxa = TRUE) %>%
  select(eventID, targetTaxonomicScope, targetLifeStageScope, hasNonTargetTaxa)
```

### nonTargetTaxa

```{r, nonTargetTaxa, message=FALSE, warning=FALSE}
humboldt_ratified <- humboldt_ratified %>%
  mutate(
    nonTargetTaxa = map_chr(eventID, ~{
      occ_names <- occ %>%
        filter(eventID == .x) %>%
        # some non target taxa were identified to higher taxa (e.g. Order = Teuthida) hence is.na(family)
        filter(!(family %in% targetTaxonomicScope[eventID == .x]) | is.na(family)) %>%
        pull(scientificName)
      paste(unique(occ_names), collapse = " | ")
    })
  )

humboldt_ratified %>% 
  filter(hasNonTargetTaxa = TRUE) %>%
  select(eventID, targetTaxonomicScope, targetLifeStageScope, hasNonTargetTaxa, nonTargetTaxa)
```

## Write data 

Write humboldt extension to file.

```{r, write-data, message=FALSE, warning=FALSE}
# Convert logical values to character strings
humboldt_ratified <- humboldt_ratified %>%
  mutate(across(where(is.logical), ~ifelse(.x, "true", "false")))

# Write the tibble to a TSV file
file_name <- paste0(format(Sys.Date(), "%Y-%m-%d"), "_humboldt.txt")
write_tsv(humboldt_ratified, here("output", file_name), na = "")
```
